## Supplementary Tables

@supptbl-cohorts distinguishes the preparation pool, original H1 inventory, and controlled analysis cohort.

::: {#supptbl-cohorts}

```{r}
#| label: cohorts-table-data
tibble::tibble(
  Stage = c(
    "Preparation and expert-screening pool",
    "Original H1 spatial model inventory",
    "Controlled spatial SPD H1 cohort"
  ),
  Records = c(
    preparation_pool_n,
    original_h1_inventory_n,
    spatial_controlled_n
  ),
  Meaning = c(
    "Records retained by the baseline pollen preparation profile",
    "Records represented in the original H1 model inventory",
    "Records eligible for time- and space-controlled spatial reporting"
  )
) |>
  kable(booktabs = TRUE)
```

**Cohorts used at different stages of the study.** Analysis-specific eligibility is determined from the registered live tables and should not be substituted for the preparation pool or original model inventory.
:::

@supptbl-pap-metrics defines the ten PAPs used in the revised article.

::: {#supptbl-pap-metrics}

```{r}
#| label: pap-metrics-table-data
table_pap_definitions |>
  kable(booktabs = TRUE)
```

**Pollen assemblage properties (PAPs).** Correlation among these summaries is expected because they describe overlapping features of the same assemblages.
:::
