## Supplementary Tables

@supptbl-cohorts distinguishes the preparation pool, sequences considered for the initial spatial analysis, and smaller analysis subsets.

::: {#supptbl-cohorts}

```{r}
#| label: cohorts-table-data
tibble::tibble(
  Stage = c(
    "Preparation and expert-screening pool",
    "Sequences considered for initial spatial analysis",
    "Available original spatial decomposition",
    "Time-controlled spatial SPD analysis",
    "Spatially controlled temporal SPD analysis",
    "Matched spatial comparison with observed events",
    "Matched temporal comparison with observed events"
  ),
  Count = c(
    preparation_pool_n,
    original_h1_inventory_n,
    spatial_available_n,
    spatial_controlled_n,
    temporal_spd_n,
    event_spatial_n,
    event_temporal_n
  ),
  Unit = c(
    "pollen sequences",
    "pollen sequences",
    "pollen sequences",
    "pollen sequences",
    "continental-group–age comparisons",
    "pollen sequences",
    "continental-group–age comparisons"
  ),
  Meaning = c(
    "Sequences retained by the baseline pollen preparation profile",
    "Sequences considered for the initial spatial analysis",
    "Original spatial models with an estimable HVarPart decomposition",
    "Sequences eligible for time-controlled spatial reporting",
    "Continental-group time slices eligible after spatial control",
    "Sequences with observed events and three estimable proxy formulations",
    "Continental-group time slices with observed events and three estimable proxy formulations"
  )
) |>
  kable(booktabs = TRUE)
```

**Subsets used at different stages of the study.** Analysis-specific eligibility is determined from the registered live tables and should not be substituted for the preparation pool or initial spatial analysis set.
:::

@supptbl-pap-metrics defines the ten PAPs used in the revised article.

::: {#supptbl-pap-metrics}

```{r}
#| label: pap-metrics-table-data
table_pap_definitions |>
  kable(booktabs = TRUE)
```

**Pollen assemblage properties (PAPs).** Correlation among these summaries is expected because they describe overlapping features of the same assemblages.
:::
