---
format:
  html:
    self-contained: true
  pdf:
    keep-tex: false
---

## Response Draft: Pollen-Sum Filtering

```{r}
#| label: setup-pollen-response
#| include: false
#| warning: false
#| message: false

repository_root <- normalizePath(getwd(), winslash = "/")
while (!file.exists(file.path(repository_root, "R", "00_Config_file.R"))) {
  repository_parent <- dirname(repository_root)
  if (identical(repository_parent, repository_root)) {
    stop("Could not locate repository root.")
  }
  repository_root <- repository_parent
}

library(dplyr)
library(readr)
library(stringr)
library(purrr)
library(knitr)

fmt_int <- function(x) {
  format(round(x), big.mark = ",", scientific = FALSE, trim = TRUE)
}

fmt_pct <- function(x, digits = 1) {
  stringr::str_c(
    format(round(100 * x, digits), nsmall = digits, trim = TRUE),
    "%"
  )
}

path_rowsum <-
  file.path(repository_root,
    "Outputs", "Tables", "Dataset_summaries", "Pollen",
    "pollen__sample_rowsums__climate_zone.csv"
  )
path_ret_overall <-
  file.path(repository_root,
    "Outputs", "Tables", "Dataset_summaries", "Pollen",
    "pollen__threshold_retention__overall.csv"
  )

required_paths <-
  c(
    path_rowsum,
    path_ret_overall
  )

missing_paths <- required_paths[!file.exists(required_paths)]
if (
  length(missing_paths) > 0
) {
  stop(
    stringr::str_c(
      "Missing required input files:",
      stringr::str_c(missing_paths, collapse = "; "),
      sep = " "
    ),
    call. = FALSE
  )
}

rowsum <-
  readr::read_csv(
    path_rowsum,
    show_col_types = FALSE
  )
ret_overall <-
  readr::read_csv(
    path_ret_overall,
    show_col_types = FALSE
  ) |>
  dplyr::arrange(.data$threshold)

baseline_threshold <-
  ret_overall |>
  dplyr::summarise(
    value = min(.data$threshold, na.rm = TRUE)
  ) |>
  dplyr::pull(.data$value)

thresholds <-
  ret_overall |>
  dplyr::distinct(.data$threshold) |>
  dplyr::arrange(.data$threshold) |>
  dplyr::pull(.data$threshold)

thresholds_nonbaseline <- thresholds[thresholds > baseline_threshold]

rowsum_count_based <-
  rowsum |>
  dplyr::filter(.data$pollen_percentage == FALSE)

rowsum_values <-
  rowsum_count_based |>
  dplyr::pull(.data$rowsum)

share_samples_ge <-
  purrr::map_dbl(
    thresholds,
    ~ mean(rowsum_values >= .x, na.rm = TRUE)
  )

names(share_samples_ge) <- as.character(thresholds)

n_samples_baseline <- nrow(rowsum_count_based)

fig_rowsum <-
  file.path(repository_root,
    "Outputs",
    "Figures",
    "Data",
    "pollen__sample_density__climate_zone.png"
  )
fig_region <-
  file.path(repository_root,
    "Outputs",
    "Figures",
    "Data",
    "pollen__threshold_retention__region.png"
  )
figure_paths <- c(
  fig_rowsum,
  fig_region
)

missing_figures <- figure_paths[!file.exists(figure_paths)]
if (
  length(missing_figures) > 0
) {
  stop(
    stringr::str_c(
      "Missing required figure files:",
      stringr::str_c(missing_figures, collapse = "; "),
      sep = " "
    ),
    call. = FALSE
  )
}
```


We appreciate the reviewer's concern regarding low pollen-count samples and potential impacts on PAP estimates.

We agree that this point requires explicit quantification and clarification in both the manuscript and response letter. We have now added targeted diagnostics that directly evaluate retained sample-count distributions and the consequences of stricter pollen-count thresholds for sequence and sample retention.

First, we clarify that the low-count values are used as filtering thresholds in preprocessing, not as a blanket forcing of all samples to one fixed count in the diagnostics presented here. To evaluate whether low-count samples dominate, we quantified the empirical pollen-count distribution among count-based samples in the baseline filtered data ($N = `r fmt_int(n_samples_baseline)`$ samples).

```{r}
#| label: text-sample-shares
#| echo: false
#| results: asis
share_lines <-
  purrr::map_chr(
    thresholds,
    ~ stringr::str_glue(
      "{fmt_pct(share_samples_ge[[as.character(.x)]])} of samples have >={fmt_int(.x)} grains"
    )
  )

cat(
  stringr::str_glue(
    "{stringr::str_c(share_lines, collapse = '; ')}.\n"
  )
)
```

Thus, the retained count-based samples are not dominated by very low pollen counts. Percentage-only sequences are excluded from this distribution because their values are not grain counts.

```{r}
#| label: fig-density-rowsum-climatezone
#| fig-cap: "Terrestrial pollen counts in count-based samples by continental group and region."
#| out-width: "100%"
#| echo: false
knitr::include_graphics(fig_rowsum)
```

Second, we evaluated sensitivity to stricter thresholds by recomputing retention at higher count criteria relative to the baseline threshold (`r fmt_int(baseline_threshold)`).

```{r}
#| label: text-retention-summary
#| results: asis
#| echo: false
ret_nonbase <-
  ret_overall |>
  dplyr::filter(
    .data$threshold %in% thresholds_nonbaseline
  ) |>
  dplyr::arrange(.data$threshold)

if (
  nrow(ret_nonbase) > 0
) {
  rec_txt <-
    ret_nonbase |>
    dplyr::transmute(
      summary = stringr::str_glue(
        "{fmt_pct(.data$prop_retained)} at {fmt_int(.data$threshold)}"
      )
    ) |>
    dplyr::pull(.data$summary) |>
    stringr::str_c(collapse = ", ")

  samp_txt <-
    ret_nonbase |>
    dplyr::transmute(
      summary = stringr::str_glue(
        "{fmt_pct(.data$prop_samples_retained)} at {fmt_int(.data$threshold)}"
      )
    ) |>
    dplyr::pull(.data$summary) |>
    stringr::str_c(collapse = ", ")

  cat(stringr::str_glue("Sequence retention is {rec_txt}.\n\n"))
  cat(stringr::str_glue("Sample retention is {samp_txt}.\n"))
} else {
  cat("No stricter-than-baseline thresholds were found in the retention table.\n")
}
```

These results show the expected trade-off: stricter thresholds reduce potential low-count bias but also remove substantial fractions of sequences and samples, particularly in data-sparse regions.

Third, we present retention by continental group to show where losses occur under stricter filtering. This makes the representational trade-off transparent rather than implicit.

In response to the reviewer, we have therefore strengthened both the methods text and supplementary evidence by adding: (i) explicit reporting of retained pollen-sum distributions, (ii) explicit retention sensitivity analyses at higher thresholds, and (iii) clearer framing that PAP-based analyses are in the pollen domain and remain subject to known compositional and production biases.

```{r}
#| label: fig-threshold-retention-region
#| fig-cap: "Sequence and sample retention under stricter pollen-count thresholds by continental group."
#| out-width: "100%"
#| echo: false
knitr::include_graphics(fig_region)
```

### Data-derived summary table

```{r}
#| label: tbl-retention-overall
#| tbl-cap: "Overall retention metrics by pollen-sum threshold."
#| echo: false
#| error: false
#| warning: false

ret_table <-
  ret_overall |>
  dplyr::select(
    "threshold",
    "n_records_retained",
    "n_samples_retained",
    "prop_retained",
    "prop_samples_retained"
  ) |>
  dplyr::mutate(
    threshold = fmt_int(.data$threshold),
    n_records_retained = fmt_int(.data$n_records_retained),
    n_samples_retained = fmt_int(.data$n_samples_retained),
    prop_retained = fmt_pct(.data$prop_retained),
    prop_samples_retained = fmt_pct(.data$prop_samples_retained)
  ) |>
  dplyr::rename(
    "Threshold" = .data$threshold,
    "Sequences retained" = .data$n_records_retained,
    "Samples retained" = .data$n_samples_retained,
    "Sequence retention" = .data$prop_retained,
    "Sample retention" = .data$prop_samples_retained
  )

knitr::kable(ret_table, align = c("r", "r", "r", "r", "r"))
```

### Notes for manuscript integration

- Methods: Add a short paragraph distinguishing threshold-based filtering decisions from downstream empirical distribution diagnostics.
- Discussion/limitations: Add a short paragraph framing threshold choice as a bias-coverage trade-off quantified by the sensitivity outputs.
