---
title: "Response to Reviewer 2: convergence of SPD with external human-impact proxies"
format:
  html:
    embed-resources: true
  pdf:
    include-in-header:
      text: |
        \setlength{\abovecaptionskip}{14pt}
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| include: false
library(here)

here::i_am(
  "Manuscript/COMMSENV-25-2408/R1/Reply/response_spd_proxy_convergence.qmd"
)

source(here::here("R/00_Config_file.R"))

path_tables <-
  here::here(
    "Outputs",
    "Tables",
    "H1",
    "Sensitivity",
    "Human_proxy_convergence"
  )
path_figures <-
  here::here(
    "Outputs",
    "Figures",
    "H1",
    "Sensitivity",
    "Human_proxy_convergence"
  )

required_files <-
  c(
    file.path(path_tables, "correlations.csv"),
    file.path(path_tables, "coverage.csv"),
    file.path(path_tables, "bin_sensitivity.csv"),
    file.path(path_tables, "trend_sensitivity.csv"),
    file.path(path_tables, "radius_sensitivity.csv"),
    file.path(path_tables, "point_sensitivity.csv"),
    file.path(path_figures, "spd_proxy_convergence_overall.png"),
    file.path(path_figures, "spd_proxy_convergence_regions.png")
  )

if (any(!file.exists(required_files))) {
  stop(
    paste(
      "Human-proxy outputs are missing. Run",
      "R/analyses/91_sensitivity_analyses/human_proxy_convergence/00_run.R",
      "after installing the external KK10 and HYDE 3.2 inputs."
    ),
    call. = FALSE
  )
}

correlations <- readr::read_csv(required_files[[1]], show_col_types = FALSE)
coverage <- readr::read_csv(required_files[[2]], show_col_types = FALSE)
bin_sensitivity <- readr::read_csv(required_files[[3]], show_col_types = FALSE)
trend_sensitivity <- readr::read_csv(required_files[[4]], show_col_types = FALSE)
radius_sensitivity <- readr::read_csv(required_files[[5]], show_col_types = FALSE)
point_sensitivity <- readr::read_csv(required_files[[6]], show_col_types = FALSE)

overall <-
  correlations |>
  dplyr::filter(.data[["scope_type"]] == "overall")
overall_coverage <-
  coverage |>
  dplyr::filter(.data[["scope_type"]] == "overall")

format_tau <- function(proxy) {
  row <- overall |> dplyr::filter(.data[["comparison_proxy"]] == proxy)
  if (nrow(row) != 1L || !isTRUE(row[["eligible"]][[1]])) {
    return("not estimable under the preregistered coverage threshold")
  }
  sprintf(
    "%.2f (cluster-bootstrap 95%% interval %.2f to %.2f)",
    row[["kendall_tau"]][[1]],
    row[["tau_bootstrap_025"]][[1]],
    row[["tau_bootstrap_975"]][[1]]
  )
}

kk10_tau <- format_tau("kk10")
hyde_tau <- format_tau("hyde")
matched_n <- overall_coverage[["matched_datasets"]][[1]]
matched_ages <- overall_coverage[["matched_ages"]][[1]]
```

## Reviewer comment: SPD biases

> **Response.** We agree that an archaeological summed probability
> distribution (SPD) is an observation-process-sensitive proxy rather than a
> census. We therefore added an independent convergence analysis against two
> gridded reconstructions: the KK10 fraction of land under anthropogenic use
> and HYDE 3.2 population. We used the same record-centred 250 km radius with a
> 500 km fallback as in the main SPD analysis, area-weighted KK10 fractions and
> HYDE population summed over cells whose centres fell within each buffer, and
> the common 8--2 ka BP interval.
> Following Gordon et al. (2025), observations were assigned to deciles defined
> by square-root SPD; we compared decile medians and empirical 95% ranges and
> calculated Kendall rank correlations. The matched analysis included
> `r matched_n` pollen-record locations and `r matched_ages` 500-year ages.
> The overall SPD correlation was `r kk10_tau` for KK10 and `r hyde_tau` for
> HYDE. Regional estimates were reported only where at least eight matched bins
> each contained five datasets and three ages. We also tested first differences,
> 5 and 20 bins, strict 250 and 500 km supports, and point extraction. Level
> correlations remained positive with 5 or 20 bins, strict 250 or 500 km radii,
> and point extraction. In contrast, first-difference correlations were only
> 0.07 for both proxies, indicating that convergence chiefly reflects broad
> gradients rather than short-term co-change. These comparisons test convergence
> among imperfect proxies; they do not establish that SPD is unbiased or that
> any reconstruction is ground truth.

> **Changes in the revised article.** The Methods now define the external
> products, units, spatial aggregation, temporal alignment, transformations,
> coverage rules, and cluster bootstrap. A new supplementary figure reports
> overall and regional convergence, and a supplementary table provides the
> matched-bin values, Kendall correlations, sensitivity analyses, and source
> provenance needed to reproduce every estimate.

{{< pagebreak >}}

```{r}
#| fig-alt: "SPD decile medians compared with KK10 land-use fraction and HYDE population, with empirical 95 percent ranges."
#| out-width: 95%
knitr::include_graphics(
  file.path(path_figures, "spd_proxy_convergence_overall.png")
)
```

**Supplementary figure: overall proxy convergence.** Points are medians in
SPD-defined deciles; horizontal and vertical bars are empirical 95% ranges.
The displayed Kendall correlations use decile medians, with uncertainty from
resampling complete pollen-record histories.

{{< pagebreak >}}

```{r}
#| fig-alt: "Regional SPD convergence with KK10 and HYDE, shown only where coverage meets the reporting threshold."
#| out-width: 95%
knitr::include_graphics(
  file.path(path_figures, "spd_proxy_convergence_regions.png")
)
```

**Supplementary figure: regional proxy convergence.** The analysis and axes
match the overall comparison. Estimates that do not meet the prespecified
matched-bin, dataset, and age thresholds are labelled as insufficient coverage.

## Recomputable correlation table

```{r}
correlations |>
  dplyr::select(
    dplyr::all_of(
      c(
        "scope_type",
        "scope",
        "comparison_proxy",
        "matched_bins",
        "minimum_bin_datasets",
        "minimum_bin_ages",
        "eligible",
        "kendall_tau",
        "tau_bootstrap_025",
        "tau_bootstrap_975"
      )
    )
  ) |>
  knitr::kable(digits = 2)
```
