---
title: "Response to Reviewer 2: sensitivity to archaeological search radius"
format:
  html:
    embed-resources: true
  pdf:
    include-in-header:
      text: |
        \setlength{\abovecaptionskip}{14pt}
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| include: false
library(here)

here::i_am(
  "Manuscript/COMMSENV-25-2408/Reply/response_spd_radius_sensitivity.qmd"
)

source(here::here("R/00_Config_file.R"))

path_spatial <-
  here::here("Outputs", "Tables", "H1", "Spatial", "SPD")
path_temporal <-
  here::here("Outputs", "Tables", "H1", "Temporal", "HVarPart")
path_spd <-
  here::here("Outputs", "Tables", "Data_preparation", "SPD")

coverage <-
  readr::read_csv(
    file.path(
      path_spd,
      "spd__radius_comparison__availability_by_geography.csv"
    ),
    show_col_types = FALSE
  )
spatial_summary <-
  readr::read_csv(
    file.path(
      path_spatial,
      "spd__radius_comparison__geographic_summary.csv"
    ),
    show_col_types = FALSE
  )
spatial_paired <-
  readr::read_csv(
    file.path(
      path_spatial,
      "spd__radius_comparison__dataset_paired.csv"
    ),
    show_col_types = FALSE
  )
temporal_summary <-
  readr::read_csv(
    file.path(
      path_temporal,
      "spd__radius_comparison__region_age_summary.csv"
    ),
    show_col_types = FALSE
  )
temporal_paired <-
  readr::read_csv(
    file.path(
      path_temporal,
      "spd__radius_comparison__region_age_paired.csv"
    ),
    show_col_types = FALSE
  )

coverage_overall <-
  coverage |>
  dplyr::filter(.data[["summary_level"]] == "overall")
spatial_overall <-
  spatial_summary |>
  dplyr::filter(.data[["summary_level"]] == "overall")
temporal_overall <-
  temporal_summary |>
  dplyr::filter(.data[["summary_level"]] == "overall")
temporal_human_change <-
  temporal_paired |>
  dplyr::filter(
    .data[["matched_estimable"]],
    is.finite(.data[["zero_allocation_human_delta_500_minus_250"]])
  ) |>
  dplyr::summarise(
    median_delta = stats::median(
      .data[["zero_allocation_human_delta_500_minus_250"]]
    ),
    delta_q25 = stats::quantile(
      .data[["zero_allocation_human_delta_500_minus_250"]],
      0.25
    ),
    delta_q75 = stats::quantile(
      .data[["zero_allocation_human_delta_500_minus_250"]],
      0.75
    )
  )
fallback_n_datasets <-
  coverage_overall$n_datasets[coverage_overall$radius_km == 250]
fallback_n_250 <-
  coverage_overall$n_available[coverage_overall$radius_km == 250]
fallback_n_500 <-
  fallback_n_datasets - fallback_n_250
fallback_n_valid <-
  coverage_overall$n_available[coverage_overall$radius_km == 500]
fallback_n_rescued <-
  fallback_n_valid - fallback_n_250

fmt3 <- function(x) format(round(x, 3), nsmall = 3, trim = TRUE)
```

## Reviewer comment

The reviewer questioned whether archaeological activity summarized within a
fixed distance of each pollen record adequately represents the spatial scale
of the vegetation record, and asked whether our conclusions are sensitive to
the use of a 250 km rather than a 500 km search radius.

## Response

We agree that this spatial choice requires an explicit sensitivity test.
Pollen records describe vegetation change from local to regional source areas,
whereas radiocarbon-dated archaeological sites are discrete and unevenly
distributed across the landscape.
A smaller archaeological search radius gives greater weight to evidence closer
to the pollen site, but in sparsely studied regions it may contain too few
radiocarbon dates to describe human activity reliably.
A larger radius increases the amount of archaeological evidence, but represents
a broader human landscape around the pollen site.
No single fixed radius can reproduce the pollen source area of every lake or
bog; our aim was therefore to test whether this unavoidable spatial choice
changes the ecological interpretation.

We recalculated the archaeological SPD for every pollen record using two strict
search radii, 250 km and 500 km, while keeping the radiocarbon data, calibration,
time range, smoothing, climate data, and statistical analyses unchanged.
We then compared the two radii for the same pollen records.
This design isolates the effect of changing the archaeological neighbourhood,
rather than confounding it with changes in the vegetation or climate data.

The primary analysis uses 250 km wherever sufficient archaeological evidence
is available and expands to 500 km only where the 250 km series is unavailable.
This choice retains the most spatially local estimate that the data support.
Of `r fallback_n_datasets` pollen records, `r fallback_n_250` have a usable
250 km series.
The 500 km fallback recovers a further `r fallback_n_rescued` records, giving
usable archaeological information for `r fallback_n_valid` records in total.
Thus, the fallback is a pragmatic compromise between spatial proximity and
geographical coverage, rather than an assumption that 250 km or 500 km is the
true pollen source area.

Figure 1 asks whether widening the archaeological neighbourhood changes the
relative allocation of positive hierarchical contributions between human
activity and climate for individual pollen records. Negative hierarchical
fractions are set to zero and the remaining positive human, climate, and
temporal components are renormalised to sum to one. The plotted balance is the
human share minus the climate share. Values below zero therefore indicate
greater relative climatic importance and values above zero greater relative
human importance. This composition is distinct from the absolute amount of
variation explained.
In panel A, each line connects the result for the same pollen record at 250 km
and 500 km; short lines therefore indicate little sensitivity to radius.
Panel B summarizes the shift within continents and climate zones.
The median change across `r spatial_overall$n_matched_estimable` comparable
pollen records was `r fmt3(spatial_overall$median_delta)` (interquartile range
`r fmt3(spatial_overall$delta_q25)` to
`r fmt3(spatial_overall$delta_q75)`).
The continent and climate-zone median shifts generally remain near zero,
although some intervals are broad because responses differ among records.
There is no consistent movement towards either stronger climatic or stronger
human influence when the radius is enlarged.

```{r}
#| label: fig-spd-radius-spatial
#| fig-cap: "Sensitivity of the zero-truncated human--climate importance balance to archaeological search radius. Negative hierarchical fractions were set to zero and the positive human, climate, and temporal components were renormalised to sum to one. (A) Each line connects the same pollen record analysed with archaeological evidence from within 250 km and 500 km. Negative values (teal) indicate a greater climate share, positive values (gold) indicate a greater human share, and white marks equal shares. (B) Median changes after increasing the radius from 250 km to 500 km are shown for continents and climate zones; points close to zero indicate little overall change."
#| out-width: 100%
knitr::include_graphics(
  here::here(
    "Outputs/Figures/H1/Spatial/SPD",
    paste0(
      "spd__radius_comparison__spatial_",
      "zero_truncated_human_climate_balance.png"
    )
  )
)
```

Figure 2 examines the same question through time.
It shows the share of positive hierarchical contribution attributed to human
predictors within each continent, with circles and squares representing the
250 km and 500 km radii.
The two trajectories largely overlap through the Holocene.
This indicates that the timing and relative magnitude of inferred human
influence are not materially altered by the larger archaeological search area.
Where the vegetation patterns showed little residual spatial structure, the
human share remains a valid estimate and is retained rather than being
treated as missing.

```{r}
#| label: fig-spd-radius-temporal-profiles
#| fig-cap: "Share of positive hierarchical contribution attributed to human predictors through time under the two archaeological search radii. Negative fractions were set to zero and the remaining positive human, climate, and spatial components were renormalised to sum to one. Circles show estimates based on archaeological evidence within 250 km and squares show estimates within 500 km. Continents are displayed as separate rows and the blue arrow indicates the direction from older to younger ages. The close overlap of the two trajectories shows that widening the archaeological neighbourhood has little effect on the inferred temporal pattern of human influence."
#| out-width: 100%
knitr::include_graphics(
  here::here(
    "Outputs/Figures/H1/Temporal/HVarPart",
    stringr::str_c(
      "spd__radius_comparison__temporal_human__",
      "zero_truncated_hierarchical_composition__space_control.png"
    )
  ),
  auto_pdf = FALSE
)
```

Figure 3 makes these temporal differences directly visible by subtracting the
250 km estimate from the 500 km estimate for each continent and time step.
Positive values indicate a larger zero-truncated human share at 500 km and
negative values indicate a smaller share.
The differences remain tightly centred on zero: the median change was
`r fmt3(temporal_human_change$median_delta)` (interquartile range
`r fmt3(temporal_human_change$delta_q25)` to
`r fmt3(temporal_human_change$delta_q75)`).
Only `r temporal_overall$n_ranking_reversals` of
`r temporal_overall$n_matched_estimable` comparable continent-by-time results
crossed from climate-dominated to human-dominated, or vice versa.
The temporal conclusion is therefore stable across the two radii, although we
retain these localized exceptions in the supporting tables.

```{r}
#| label: fig-spd-radius-temporal-changes
#| fig-cap: "Change in the zero-truncated human share after increasing the archaeological search radius from 250 km to 500 km. Positive values indicate a larger human share at 500 km, negative values indicate a smaller share, and zero indicates no change in the relative composition. Continents are displayed as separate rows and the blue arrow indicates the direction from older to younger ages. Values clustered around zero show that the temporal pattern is only weakly sensitive to search radius."
#| out-width: 100%
knitr::include_graphics(
  here::here(
    "Outputs/Figures/H1/Temporal/HVarPart",
    stringr::str_c(
      "spd__radius_comparison__temporal_human__",
      "zero_truncated_hierarchical_composition_change__space_control.png"
    )
  ),
  auto_pdf = FALSE
)
```

At the level of individual pollen records,
`r spatial_overall$n_ranking_reversals` of
`r spatial_overall$n_matched_estimable` comparisons crossed the boundary
between climate and human dominance.
Such a reversal can arise from a small numerical change when the original
estimate lies close to equal climate and human shares.
Importantly, these local reversals did not translate into a consistent shift
at continental or climatic-zone scales, and the temporal analysis contained
only the `r temporal_overall$n_ranking_reversals` exceptions reported above.

Together, the three figures show that changing from a 250 km to a 500 km
archaeological neighbourhood has only a small effect on the relative
composition and history of inferred human importance.
The sensitivity analysis therefore supports our methodology: using the 250 km
SPD where possible preserves the closest available link between archaeological
activity and the pollen record, while the 500 km fallback improves coverage in
archaeologically sparse areas without changing the broad ecological patterns.
Most importantly, the main large-scale result remains unchanged: climate is a
stronger predictor than human activity of broad-scale vegetation patterns.

## Methods insertion

To test whether the spatial scale used to summarize archaeological evidence
affected our conclusions, we recalculated the radiocarbon SPD for every pollen
record using strict 250 km and 500 km search radii.
Both calculations used the same radiocarbon dataset, calibration choices,
0--12 ka time range, 100-year smoothing, and minimum requirement of more than
50 radiocarbon dates.
All vegetation, climate, and analytical settings were held constant, so the
only difference between the two analyses was the archaeological search radius.
We compared results for the same pollen records and for the same continental
time steps at both radii.
The primary analysis used the 250 km SPD whenever sufficient dates were
available and the 500 km SPD otherwise, thereby favouring the most local
archaeological context while retaining records from data-sparse regions.

## Results insertion

Usable archaeological SPDs were available for
`r coverage_overall$n_available[coverage_overall$radius_km == 250]` of
`r coverage_overall$n_datasets[coverage_overall$radius_km == 250]` pollen
records at 250 km and
`r coverage_overall$n_available[coverage_overall$radius_km == 500]` at 500 km.
For the `r spatial_overall$n_matched_estimable` pollen records that could be
compared at both radii, expanding the archaeological neighbourhood produced a
median change of only `r fmt3(spatial_overall$median_delta)` in the
zero-truncated human-minus-climate importance balance (IQR
`r fmt3(spatial_overall$delta_q25)` to
`r fmt3(spatial_overall$delta_q75)`).
Although `r spatial_overall$n_ranking_reversals` individual records crossed
the boundary between climate and human dominance, the continent- and
continent- and climate-zone medians generally remained near zero, although
some subgroup intervals were broad; they showed no consistent directional
shift.
Across `r temporal_overall$n_matched_estimable` comparable continent-by-time
results, the median change in the zero-truncated human share was
`r fmt3(temporal_human_change$median_delta)` (IQR
`r fmt3(temporal_human_change$delta_q25)` to
`r fmt3(temporal_human_change$delta_q75)`), with only
`r temporal_overall$n_ranking_reversals` reversals in the relative importance
of climate and human activity.
The broad spatial and temporal patterns were therefore insensitive to the
tested archaeological search radius.

## Discussion insertion

The archaeological radius represents a trade-off between spatial proximity to
the pollen record and the amount of available evidence.
Neither tested radius should be interpreted as the pollen source area, which
varies among sites according to basin size, pollen dispersal, and the
surrounding vegetation.
Nevertheless, expanding the archaeological neighbourhood from 250 km to
500 km produced a small overall median change in the zero-truncated
human--climate importance balance, alongside heterogeneous local responses,
and did not alter its broad spatial or temporal structure.
We therefore retain the 250 km radius wherever it provides sufficient
radiocarbon evidence and use 500 km only as a fallback in data-sparse settings.
This approach preserves the most local archaeological context supported by the
data while maximizing geographical coverage.
The sensitivity analysis confirms that the main large-scale inference---that
climate is a stronger predictor than human activity of broad-scale vegetation
patterns---is robust to this choice.
