---
title: "Additional HVarPart sensitivity analysis"
format:
  html:
    embed-resources: true
  pdf: default
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| include: false
library(here)
here::i_am(
  "Manuscript/COMMSENV-25-2408/Reply/response_hvarpart_relative_importance.qmd"
)
source(here::here("R/00_Config_file.R"))

table_dir <-
  here::here("Outputs", "Tables", "Diagnostics", "HVarPart")
spatial_figure_dir <-
  here::here("Outputs", "Figures", "H1", "Spatial", "SPD")

temporal_figure_dir <-
  here::here("Outputs", "Figures", "H1", "Temporal", "HVarPart")

interrelationships_figure_dir <-
  here::here("Outputs", "Figures", "H2", "Interrelationships")
figure_dir <- here::here(
  "Outputs", "Figures", "Diagnostics", "HVarPart"
)

audit <- readr::read_csv(
  file.path(table_dir, "hvarpart__model_audit.csv"),
  show_col_types = FALSE,
  guess_max = Inf
)
profiles <- readr::read_csv(
  file.path(table_dir, "hvarpart__profile_comparison.csv"),
  show_col_types = FALSE,
  guess_max = Inf
)
audit_overall <- audit |>
  dplyr::filter(.data[["aggregation_level"]] == "analysis")
profiles_overall <- profiles |>
  dplyr::filter(.data[["aggregation_level"]] == "analysis")
spatial_spd <- profiles_overall |>
  dplyr::filter(.data[["analysis"]] == "spatial_spd")

value_for <- function(data, selected_profile, selected_predictor) {
  data |>
    dplyr::filter(
      .data[["profile"]] == selected_profile,
      .data[["predictor"]] == selected_predictor
    ) |>
    dplyr::pull(.data[["pooled_allocation"]])
}

fmt_pct <- function(x) scales::percent(x, accuracy = 0.1)
all_conclusions_stable <-
  profiles_overall |>
  dplyr::select(
    .data[["analysis"]],
    .data[["profile"]],
    .data[["predictor"]],
    .data[["pooled_allocation"]]
  ) |>
  tidyr::pivot_wider(
    names_from = "predictor",
    values_from = "pooled_allocation"
  ) |>
  dplyr::summarise(stable = all(.data[["climate"]] > .data[["human"]])) |>
  dplyr::pull(.data[["stable"]])
```

## Additional analysis and correction

During additional quality-control checks of the hierarchical variation
partitioning results, we identified an error in our previous post-processing:
negative adjusted contributions were not retained. This issue was identified
by the authors rather than raised by the reviewers. We replaced the previous
pathway with a single extractor that preserves HVarPart's `Unique`,
`Average.share`, `Individual`, `I.perc(%)`, and total adjusted $R^2$ values.

We compared three reporting profiles:

1. **Signed:** retain each unmodified individual contribution.
2. **Zero-truncated:** replace negative individual contributions with exactly
   zero and recompute the denominator from the truncated contributions.
3. **Exclude-negative:** exclude the complete model if either predictor has a
   negative individual contribution.

Within reporting group $g$, the signed allocation of predictor $p$ is

$$
A_{p,g} = \frac{\sum_m I_{p,m}}{\sum_m R^2_{\mathrm{adj},m}},
$$

where $I_{p,m}$ is HVarPart's unmodified individual contribution. Models are
otherwise excluded only when required predictors are absent, the denominator
is non-finite or non-positive, or an individual contribution is non-finite.
Every exclusion is recorded in
`Outputs/Tables/Diagnostics/HVarPart/hvarpart__model_audit.csv`.

A negative signed allocation is an adjusted partitioning result. It is not
evidence of a negative ecological effect.

## Results and presentation

For spatial SPD models, the zero-truncated allocations underlying the main
balance figure are
`r fmt_pct(value_for(spatial_spd, "zero_truncated", "climate"))` for climate
and
`r fmt_pct(value_for(spatial_spd, "zero_truncated", "human"))` for human
impact. The corresponding unmodified signed allocations are
`r fmt_pct(value_for(spatial_spd, "signed", "climate"))` and
`r fmt_pct(value_for(spatial_spd, "signed", "human"))`. Excluding complete
models with any negative contribution gives
`r fmt_pct(value_for(spatial_spd, "exclude_negative", "climate"))` for climate.

Across all five analyses, climate remains larger than human impact under all
three profiles:
`r ifelse(all_conclusions_stable, "yes", "no")`.

Because the profiles support the same overall interpretation, the main figures
use the bounded difference between paired zero-truncated human and climate
allocations. Figure 2 applies this balance spatially, Figure 3 through time,
and Figure 4 within its importance insets. Signed versions of Figures 2--4 are
provided in the Supplementary Information, and the
exclude-negative analysis is reported as an additional sensitivity profile.

```{r}
#| label: tbl-hvarpart-audit
#| tbl-cap: "Model-level HVarPart eligibility audit. Counts are not duplicated across predictor rows."
audit_overall |>
  dplyr::select(
    .data[["analysis"]],
    .data[["n_models"]],
    .data[["n_eligible"]],
    .data[["n_excluded"]],
    .data[["n_negative_unique"]],
    .data[["n_negative_individual"]]
  ) |>
  dplyr::mutate(
    analysis = stringr::str_replace_all(.data[["analysis"]], "_", " ")
  ) |>
  knitr::kable(
    col.names = c(
      "Analysis", "Models", "Eligible", "Excluded",
      "Neg. unique", "Neg. individual"
    )
  )
```

```{r}
#| label: fig-hvarpart-profiles
#| fig-cap: "Zero-truncated main-figure allocation and two sensitivity profiles, calculated from the preserved fitted HVarPart results."
#| out-width: "100%"
knitr::include_graphics(
  file.path(figure_dir, "hvarpart_profile_comparison.png")
)
```

## Figure changes

Figure 2 now expresses the complementary human and climate allocations as
human minus climate importance. The bounded scale runs from -1 (climate
impact) through 0 (equal importance) to 1 (human impact). For each continent,
the figure combines the continent-wide distribution of record-level balances,
pooled climate-zone balances, and a horizontal line for the pooled continental
balance. Densities use a continuous climate-to-neutral-to-human colour scale;
the climate-zone panels use the same scale as a pale background while their
summary points retain categorical climate-zone colours. Maps and continental
density panels do not use the background gradient. The supplementary
counterpart reports the untruncated human-minus-climate hierarchical
contribution difference with the same spatially filtered summaries.

```{r}
#| label: embed-hvarpart-spatial-main
#| out-width: "100%"
knitr::include_graphics(
  file.path(
    spatial_figure_dir,
    stringr::str_c(
      "spd",
      "human_climate_balance",
      "zero_truncated_hierarchical_composition",
      "time_and_space_control.png",
      sep = "__"
    )
  )
)
```

*Revised main manuscript Figure 2. Spatial balance between zero-truncated human and climate allocations.*

```{r}
#| label: embed-hvarpart-spatial-signed
#| out-width: "100%"
knitr::include_graphics(
  file.path(
    spatial_figure_dir,
    stringr::str_c(
      "spd",
      "human_climate_balance",
      "untruncated_hierarchical_contribution_difference",
      "time_and_space_control.png",
      sep = "__"
    )
  )
)
```

*Supplementary signed Figure 2. Untruncated human-minus-climate hierarchical contribution difference with time and space control.*

The main temporal H1 figure shows the balance through time as lollipops drawn
from equal importance to each estimate. Point shape distinguishes the SPD and
event human-impact proxies, while point colour uses the same
climate-neutral-human scale as Figure 2 against a pale version of that
gradient. SPD estimates are restricted to 2--8.5 ka BP, whereas event-based
estimates span 0.5--8.5 ka BP. Its supplementary counterpart applies the same
age domains and shows the unmodified signed allocations as grouped, unstacked
bars with zero and one reference lines. The H2 main-figure insets
likewise use one outlined balance lollipop per region and climate zone against
the pale gradient; supplementary insets retain separate unmodified signed
human and climate allocations.

```{r}
#| label: embed-hvarpart-temporal-main
#| out-width: "100%"
knitr::include_graphics(
  file.path(
    temporal_figure_dir,
    stringr::str_c(
      "spd_events__human_climate_space__",
      "zero_truncated_hierarchical_composition__space_control.png"
    )
  )
)
```

*Revised main manuscript Figure 3. Temporal balance between zero-truncated human and climate allocations.*

```{r}
#| label: embed-hvarpart-temporal-signed
#| out-width: "100%"
knitr::include_graphics(
  file.path(
    temporal_figure_dir,
    stringr::str_c(
      "human_climate_only_",
      "untruncated_hierarchical_contributions.png"
    )
  )
)
```

*Supplementary signed Figure 3. Unmodified signed climate and human allocations through time.*

```{r}
#| label: embed-hvarpart-h2-main
#| out-width: "100%"
knitr::include_graphics(
  file.path(
    interrelationships_figure_dir,
    "predictor_interrelationships.png"
  )
)
```

*Revised main manuscript Figure 4. Multidimensional community-change ordinations with balance lollipops for human and climate importance.*

```{r}
#| label: embed-hvarpart-h2-signed
#| out-width: "100%"
knitr::include_graphics(
  file.path(
    interrelationships_figure_dir,
    stringr::str_c(
      "predictor_interrelationships_",
      "untruncated_hierarchical_contributions.png"
    )
  )
)
```

*Supplementary signed Figure 4. Multidimensional community-change ordinations with separate unmodified signed human and climate allocations.*
