---
title: "Explained variation and human importance"
format:
  html:
    embed-resources: true
  pdf: default
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| include: false
library(here)
here::i_am(
  paste0(
    "Manuscript/COMMSENV-25-2408/Reply/",
    "response_hvarpart_variance_explained.qmd"
  )
)
source(here::here("R/00_Config_file.R"))
table_dir <- here::here("Outputs", "Tables", "HVarPart")
figure_dir <- here::here(
  "Outputs", "Figures", "Diagnostics", "HVarPart"
)
audit <- readr::read_csv(
  file.path(table_dir, "hvarpart_variance_missing_negative_audit.csv"),
  show_col_types = FALSE
)
distribution_records <- readr::read_csv(
  file.path(
    table_dir,
    "HVarPart_spatial_adjr2_distribution_records.csv"
  ),
  show_col_types = FALSE
)
distribution_summary <-
  distribution_records |>
  dplyr::summarise(
    n_models = dplyr::n(),
    mean = mean(.data[["adjusted_r_squared"]]),
    q1 = stats::quantile(
      .data[["adjusted_r_squared"]],
      probs = 0.25,
      names = FALSE
    ),
    median = stats::median(.data[["adjusted_r_squared"]]),
    q3 = stats::quantile(
      .data[["adjusted_r_squared"]],
      probs = 0.75,
      names = FALSE
    )
  )
signed <- readr::read_csv(
  file.path(
    table_dir,
    "HVarPart_adjr2_human_importance_overall_statistics.csv"
  ),
  show_col_types = FALSE
)
```

## Response to the reviewer

We thank the reviewer for asking how much variation the human-impact and
climate predictors explain, and whether model fit covaries with the estimated
importance of human impact.

We added a direct commonality decomposition for every available H1 model.
Unique human, unique climate, shared, and total adjusted R² fractions come
from `Var.part`.
Hierarchical individual importance remains a separate quantity from
`Hier.part`.
This follows the
[rdacca.hp contract](https://search.r-project.org/CRAN/refmans/rdacca.hp/html/rdacca.hp.html)
and [Lai et al. (2022)](https://doi.org/10.1111/2041-210X.13800).

For each model, total adjusted R² equals unique human plus unique climate plus
shared explained variation.
Adding unexplained variation reconstructs one.
Signed adjusted fractions are primary; the complete bounded decomposition
sensitivity remains available in the source tables.
These are joint multivariate models of ten pollen-derived properties, not ten
independent fits.

Across the four H1 analyses, `r sum(audit[["n_available"]])` of
`r sum(audit[["n_models"]])` model records had complete decompositions.
No available model exceeded the 0.001 accounting tolerance.
For spatial SPD, `r distribution_summary[["n_models"]]` decompositions had a
median adjusted R² of
`r formatC(distribution_summary[["median"]], digits = 3, format = "f")`
(interquartile range
`r formatC(distribution_summary[["q1"]], digits = 3, format = "f")` to
`r formatC(distribution_summary[["q3"]], digits = 3, format = "f")`).
We show record-level distributions rather than sums because sums scale
directly with the number of models in each group.

```{r}
#| label: tbl-hvarpart-variance-audit
#| tbl-cap: "Availability, accounting, and signed-component audit for H1."
audit |>
  dplyr::select(
    .data[["analysis"]],
    .data[["n_models"]],
    .data[["n_available"]],
    .data[["n_accounting_invalid"]],
    .data[["n_negative_unique_human"]],
    .data[["n_negative_unique_climate"]],
    .data[["n_negative_shared"]]
  ) |>
  knitr::kable(
    col.names = c(
      "Analysis", "Models", "Available", "Invalid",
      "Neg. human", "Neg. climate", "Neg. shared"
    )
  )
```

## Three independent figure products

The first figure shows the distribution of signed adjusted R² values within
each continent and climate-zone combination.
Points are individual spatial SPD models, boxes show medians and interquartile
ranges, and violins show distribution shape when at least three varying values
are available.
Finite negative adjusted R² values are retained.

```{r}
#| label: fig-adjr2-distribution
#| fig-cap: "Adjusted R-squared distributions by continent and climate zone. Points are individual models; boxes show medians and interquartile ranges."
#| out-width: "100%"
#| fig-pos: H
knitr::include_graphics(
  file.path(figure_dir, "HVarPart_spatial_adjr2_distribution.png")
)
```

The correlation figures use signed relative human importance only; no
zero-truncated importance figure is shown.
Across `r signed[["n_models"]]` eligible spatial SPD models, Spearman ρ =
`r formatC(signed[["spearman_rho"]], digits = 3, format = "f")`
and Pearson r =
`r formatC(signed[["pearson_r"]], digits = 3, format = "f")`.

```{r}
#| label: fig-overall-adjr2-importance
#| fig-cap: "Adjusted R-squared and signed human relative importance across eligible spatial SPD models."
#| out-width: "75%"
#| fig-pos: H
knitr::include_graphics(
  file.path(
    figure_dir,
    "HVarPart_adjr2_human_importance_overall.png"
  )
)
```

The grid uses five continent rows and eleven climate-zone columns.
Populated panels report n and Spearman ρ.
Statistics and trend lines require at least three observations and non-zero
variation.
Absent combinations remain explicit empty cells.

```{r}
#| label: fig-grid-adjr2-importance
#| fig-cap: "Adjusted R-squared and signed human relative importance by continent and climate zone."
#| out-width: "100%"
#| fig-pos: H
knitr::include_graphics(
  file.path(figure_dir, "HVarPart_adjr2_human_importance_grid.png")
)
```

All model values, distribution summaries, audits, and signed correlation
statistics are exported under `Outputs/Tables/HVarPart`.
The validation script independently reconstructs every distribution summary,
model count, Spearman coefficient, and Pearson coefficient.
