---
title: "Reciprocal spatial and temporal control of the H1 analysis"
format:
  html:
    embed-resources: true
  pdf: default
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| include: false
library(here)
here::i_am(
  "Manuscript/COMMSENV-25-2408/Reply/response_spatiotemporal_control.qmd"
)
source(here::here("R/00_Config_file.R"))

table_dir <-
  here::here(
    "Outputs",
    "Tables",
    "Diagnostics",
    "Spatiotemporal_dependence"
  )
spatial_figure_dir <- here::here(
  "Outputs", "Figures", "H1", "Spatial", "SPD"
)

temporal_figure_dir <- here::here(
  "Outputs", "Figures", "H1", "Temporal", "HVarPart"
)
robustness <- readr::read_csv(
  file.path(
    table_dir,
    stringr::str_c(
      "spd__human_climate_balance__robustness__",
      "time_and_space_control.csv"
    )
  ),
  show_col_types = FALSE
)
human_climate_only <- readr::read_csv(
  file.path(
    table_dir,
    "human_climate_balance__matched_estimates__human_climate_only.csv"
  ),
  show_col_types = FALSE
)
controlled <- readr::read_csv(
  file.path(
    table_dir,
    stringr::str_c(
      "spd__human_climate_balance__estimates__",
      "time_and_space_control.csv"
    )
  ),
  show_col_types = FALSE
)
time_status <- readr::read_csv(
  file.path(
    table_dir,
    "spd_events__human_climate_time__status__time_control.csv"
  ),
  show_col_types = FALSE
)
temporal_moran <- readr::read_csv(
  file.path(
    table_dir,
    "spd_events__human_climate_time__residual_moran__time_control.csv"
  ),
  show_col_types = FALSE
)
space_status <- readr::read_csv(
  file.path(
    table_dir,
    "spd_events__human_climate_space__status__space_control.csv"
  ),
  show_col_types = FALSE
)
spatial_moran <- readr::read_csv(
  file.path(
    table_dir,
    "spd_events__human_climate_space__residual_moran__space_control.csv"
  ),
  show_col_types = FALSE
)
rankings <- readr::read_csv(
  file.path(
    table_dir,
    "spd_events__human_climate_space__rankings__space_control.csv"
  ),
  show_col_types = FALSE
)
balance_summary <-
  human_climate_only |>
  dplyr::filter(.data[["aggregation_level"]] == "overall") |>
  dplyr::select(
    .data[["profile"]],
    original_balance = .data[["importance_balance"]]
  ) |>
  dplyr::left_join(
    controlled |>
      dplyr::filter(.data[["aggregation_level"]] == "overall") |>
      dplyr::select(
        .data[["profile"]],
        controlled_balance = .data[["adjusted_balance"]]
      ),
    by = "profile"
  ) |>
  dplyr::left_join(robustness, by = "profile")
diagnostic_summary <-
  tibble::tibble(
    diagnostic = c(
      "Estimable time-controlled core-proxy models",
      "Models with residual temporal dependence",
      "Region-age models with selected spatial terms",
      "Models with residual spatial dependence",
      "Estimable human-climate ranking changes"
    ),
    count = c(
      sum(stringr::str_starts(time_status[["status"]], "estimated")),
      temporal_moran |>
        dplyr::filter(
          .data[["stage"]] == "time_controlled",
          .data[["positive_autocorrelation"]]
        ) |>
        dplyr::distinct(
          .data[["analysis"]],
          .data[["dataset_id"]]
        ) |>
        nrow(),
      sum(space_status[["status"]] == "spatial_model_estimated"),
      spatial_moran |>
        dplyr::filter(.data[["positive_autocorrelation"]]) |>
        dplyr::distinct(
          .data[["analysis"]],
          .data[["region"]],
          .data[["age"]]
        ) |>
        nrow(),
      sum(rankings[["ranking_changed"]], na.rm = TRUE)
    ),
    denominator = c(
      nrow(time_status),
      sum(stringr::str_starts(time_status[["status"]], "estimated")),
      nrow(space_status),
      nrow(space_status),
      sum(!is.na(rankings[["ranking_changed"]]))
    )
  )
```

## Response to the reviewer

We agree that the H1 comparison must address both dimensions of dependence:
repeated ages within a pollen core and nearby cores within a time slice. We
therefore expanded the sensitivity analysis into reciprocal temporal and
spatial control while retaining human impact and climate as the substantive
drivers. Time and space are structural controls, not equivalent causal
competitors.

### Figure 2: time within cores, space during aggregation

We first enforced one analytical row per core and 500-year age. Repeated PAP
response estimates at the same age were averaged only after verifying that
predictors were invariant. Conflicts and missing ages cause explicit failures.
The human-plus-climate-only model is retained as a reference. The controlled
model adds centred and scaled linear age as a third group. Linear age is used
because cores contain at most 14 unique ages; temporal MEMs would consume too
many residual degrees of freedom and require a separate detrending decision.

The main Figure 2 value remains one bounded number per core:

$$
\frac{\max(I_H,0)-\max(I_C,0)}
     {\max(I_H,0)+\max(I_C,0)}.
$$

Cores for which both contributions are non-positive are reported and excluded
from this presentation profile. Continent-specific positive dbMEMs are selected
from the signed time-controlled human-minus-climate difference conditional on
continent-by-climate-zone strata and reused for the bounded balance. Markers
are predictions at dbMEM equal to zero; distributions show observed controlled
core values.

```{r}
knitr::kable(
  balance_summary |>
    dplyr::select(
      .data[["profile"]],
      .data[["original_balance"]],
      .data[["controlled_balance"]],
      .data[["min_thinning_agreement"]],
      .data[["robustness_class"]]
    ),
  digits = 3
)
```

```{r}
#| fig-cap: >-
#|   Time-controlled core balances and spatially filtered regional markers.
knitr::include_graphics(
  file.path(
    spatial_figure_dir,
    stringr::str_c(
      "spd",
      "human_climate_balance",
      "zero_truncated_hierarchical_composition",
      "time_and_space_control.png",
      sep = "__"
    )
  )
)
```

### Figure 3: space within time slices

Each region-by-age model uses one row per core. Positive dbMEMs are selected
against the multivariate PAP response conditional on the complete human and
climate groups using a global test and Blanchet double stopping. When no
conditional spatial signal is detected, human-climate-only contributions
contributions are retained and space is set to exactly zero for the
presentation stack. Non-estimable models remain missing.

The revised main Figure 3 uses zero-truncated hierarchical contributions for
human impact, climate, and space. Signed hierarchical contributions and pure
partial adjusted-R-squared fractions are exported and plotted separately; they
are related diagnostics, not interchangeable estimates.

```{r}
#| fig-cap: "Human, climate, and selected spatial contributions through time."
knitr::include_graphics(
  file.path(
    temporal_figure_dir,
    stringr::str_c(
      "spd_events__human_climate_space__",
      "zero_truncated_hierarchical_composition__space_control.png"
    )
  )
)
```

```{r}
knitr::kable(diagnostic_summary)
```

Residual temporal or spatial dependence remains visible in the diagnostics
rather than being treated as automatically removed. The original two-group
analyses, signed contributions, pure fractions, actual permutation counts,
selection details, thinning replicates, exclusions, and provenance are retained
as supplementary machine-readable outputs. Negative adjusted fractions are
never replaced with arbitrary positive values.

## Interpretation boundary

The analysis tests whether the human-versus-climate conclusion is stable under
explicit structural controls. It does not establish that time or space causes
the PAP response, and it does not resolve finer-scale temporal autocorrelation
within the original irregular pollen levels beyond the 500-year analysis units.
