---
title: "Spatial-dependence sensitivity of the H1 analysis"
format:
  html:
    embed-resources: true
  pdf: default
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| include: false
library(here)
here::i_am(
  "Manuscript/COMMSENV-25-2408/Reply/response_spatial_dependence.qmd"
)
source(here::here("R/00_Config_file.R"))

table_dir <-
  here::here(
    "Outputs",
    "Tables",
    "Diagnostics",
    "Spatiotemporal_dependence"
  )
figure_dir <-
  here::here(
    "Outputs", "Figures", "H1", "Spatial", "SPD",
    "Spatial_dependence"
  )

composition_figure_dir <-
  here::here(
    "Outputs", "Figures", "H1", "Spatial", "SPD",
    "Spatiotemporal_composition"
  )
robustness <- readr::read_csv(
  file.path(
    table_dir,
    "spd__human_climate_balance__robustness__space_control.csv"
  ),
  show_col_types = FALSE
)
sensitivity <- readr::read_csv(
  file.path(
    table_dir,
    "spd__human_climate_balance__sensitivity__space_control.csv"
  ),
  show_col_types = FALSE
)
estimates <- readr::read_csv(
  file.path(
    table_dir,
    "spd__human_climate_balance__estimates__space_control.csv"
  ),
  show_col_types = FALSE
)
status <- readr::read_csv(
  file.path(
    table_dir,
    "spd__human_climate_space__status__space_control.csv"
  ),
  show_col_types = FALSE
)
partial <- readr::read_csv(
  file.path(
    table_dir,
    "spd__human_climate_space__unique_adjusted_r2__space_control.csv"
  ),
  show_col_types = FALSE
)
moran <- readr::read_csv(
  file.path(
    table_dir,
    "spd__human_climate_balance__moran_diagnostics__space_control.csv"
  ),
  show_col_types = FALSE
)
residual_moran <- readr::read_csv(
  file.path(
    table_dir,
    "spd__human_climate_space__residual_moran__space_control.csv"
  ),
  show_col_types = FALSE
)
baseline_signed <-
  sensitivity |>
  dplyr::filter(
    .data[["sensitivity_type"]] == "unthinned",
    .data[["aggregation_level"]] == "overall",
    .data[["profile"]] == "signed"
  ) |>
  dplyr::pull("importance_balance")
adjusted_signed <-
  estimates |>
  dplyr::filter(
    .data[["aggregation_level"]] == "overall",
    .data[["profile"]] == "signed"
  ) |>
  dplyr::pull("adjusted_balance")
thinning_agreement <-
  sensitivity |>
  dplyr::filter(
    .data[["sensitivity_type"]] == "thinning",
    .data[["aggregation_level"]] == "overall",
    .data[["profile"]] == "signed"
  ) |>
  dplyr::group_by(.data[["distance_km"]]) |>
  dplyr::summarise(
    climate_agreement = mean(.data[["ranking"]] == "climate"),
    .groups = "drop"
  )
agreement_250 <-
  thinning_agreement[["climate_agreement"]][
    thinning_agreement[["distance_km"]] == 250
  ]
agreement_500 <-
  thinning_agreement[["climate_agreement"]][
    thinning_agreement[["distance_km"]] == 500
  ]
n_spatial_models <- sum(status[["status"]] == "spatial_model_estimated")
n_temporal_models <- nrow(status)
n_residual_groups <-
  residual_moran |>
  dplyr::filter(.data[["positive_autocorrelation"]]) |>
  dplyr::distinct(.data[["region"]], .data[["age"]]) |>
  nrow()
fmt_num <- function(value) format(round(value, 3), nsmall = 3)
fmt_pct <- function(value) scales::percent(value, accuracy = 1)
```

## Response to the reviewer

We agree that spatial dependence among pollen records can make densely sampled
areas disproportionately influential. We therefore added a reproducible
sensitivity analysis that combines spatial subsampling with explicit spatial
control.

The models contributing to Figure 2 are fitted separately within pollen cores.
Coordinates are constant within a core and therefore cannot be included as a
within-core predictor. We instead modelled spatial structure in the resulting
core-level human-minus-climate importance balances. Positive distance-based
Moran eigenvector maps (dbMEMs) were constructed independently within
continents and selected conditional on the existing continent-by-climate-zone
strata. We diagnosed Moran's I before and after filtering at 250 km and 500 km.

As a design-based sensitivity, we also repeated the Figure 2 aggregation after
100 random within-stratum thinnings at minimum distances of 250 km and 500 km.
We additionally omitted each continent and each climate zone in turn. The
classification was specified in advance: robust conclusions require agreement
of the spatially filtered and deterministic estimates plus at least 95%
agreement among thinning replicates.

The signed overall balance changed only from
`r fmt_num(baseline_signed)` to `r fmt_num(adjusted_signed)` after dbMEM
filtering. Climate remained the larger contribution in
`r fmt_pct(agreement_250)`
of 250-km replicates and
`r fmt_pct(agreement_500)`
of 500-km replicates, and in every deterministic leave-out analysis. Both
signed and zero-truncated profiles therefore meet the predeclared **robust**
ranking criterion.

```{r}
knitr::kable(
  robustness |>
    dplyr::transmute(
      Profile = .data[["profile"]],
      Classification = .data[["robustness_class"]],
      `Minimum thinning agreement` =
        scales::percent(
          .data[["min_thinning_agreement"]],
          accuracy = 1
        ),
      `Residual dependence detected` =
        .data[["residual_spatial_dependence"]]
    )
)
```

```{r}
#| fig-cap: "Spatial autocorrelation before and after dbMEM filtering."
knitr::include_graphics(
  file.path(figure_dir, "spd__human_climate_balance__moran.png")
)
```

```{r}
#| fig-cap: "Human-minus-climate balance after repeated spatial thinning."
knitr::include_graphics(
  file.path(figure_dir, "spd__human_climate_balance__spatial_thinning.png")
)
```

For the region-by-age analysis, sites occur together within each model, so
space can be included directly. We selected dbMEMs conditional on the complete
human and climate predictor sets, fitted a three-group hierarchical partition,
and separately calculated pure human, pure climate, and pure spatial adjusted
$R^2$ fractions. Hierarchical contributions and pure conditional fractions are
reported separately because they answer different questions.

Spatial terms were selected in `r n_spatial_models` of
`r n_temporal_models` region-by-age models. Detectable positive residual
autocorrelation remained on at least one diagnostic axis and distance scale in
`r n_residual_groups` groups. Likewise, the Figure 2 residuals retained
short-range autocorrelation at 250 km, although not at 500 km. We therefore
interpret the ranking as robust to the implemented spatial controls, but do not
claim that the data or residuals are spatially independent.

```{r}
#| fig-cap: "Pure human, climate, and spatial fractions through time."
knitr::include_graphics(
  file.path(
    composition_figure_dir,
    "spd__human_climate_time__unique_adjusted_r2.png"
  )
)
```

Residual spatial dependence is reported rather than treated as automatically
removed. Any region-age group with insufficient locations, rank limitations,
or detectable residual autocorrelation remains explicitly flagged in the
exported diagnostics.
