---
title: "Response: HVarPart relative importance"
format:
  html:
    embed-resources: true
  pdf: default
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(here)
library(dplyr)
library(readr)
library(tidyr)

table_dir <- here::here("Outputs", "Tables", "HVarPart")
figure_dir <- here::here(
  "Outputs", "Figures", "Extended_data_figures", "HVarPart"
)

audit <- read_csv(
  file.path(table_dir, "hvarpart_model_audit.csv"),
  show_col_types = FALSE,
  guess_max = Inf
)
profiles <- read_csv(
  file.path(table_dir, "hvarpart_profile_comparison.csv"),
  show_col_types = FALSE,
  guess_max = Inf
)
validation <- read_csv(
  file.path(table_dir, "hvarpart_validation_provenance.csv"),
  show_col_types = FALSE
)

audit_overall <- audit |>
  filter(aggregation_level == "analysis")
profiles_overall <- profiles |>
  filter(aggregation_level == "analysis")
spatial_spd <- profiles_overall |>
  filter(analysis == "spatial_spd")

value_for <- function(data, selected_profile, selected_predictor) {
  data |>
    filter(
      profile == selected_profile,
      predictor == selected_predictor
    ) |>
    pull(pooled_allocation)
}

fmt_pct <- function(x) scales::percent(x, accuracy = 0.1)
all_conclusions_stable <- profiles_overall |>
  select(analysis, profile, predictor, pooled_allocation) |>
  pivot_wider(names_from = predictor, values_from = pooled_allocation) |>
  summarise(stable = all(climate > human)) |>
  pull(stable)
```

## Response

We thank the reviewer for identifying that the previous post-processing did
not preserve negative adjusted hierarchical-partitioning contributions. We
have replaced that pathway with a single extractor that retains HVarPart's
`Unique`, `Average.share`, `Individual`, `I.perc(%)`, and total adjusted

$R^2$ values without truncation.

Our primary estimand is now the **signed allocation of adjusted explained
variation**. Within reporting group $g$, predictor $p$ is pooled as

$$
A_{p,g} = \frac{\sum_m I_{p,m}}{\sum_m R^2_{\mathrm{adj},m}},
$$

where $I_{p,m}$ is HVarPart's unmodified individual contribution. Models are
excluded only when required predictors are absent, the total adjusted $R^2$ is
non-finite or non-positive, or individual contributions are non-finite. Every
exclusion is recorded in
`Outputs/Tables/HVarPart/hvarpart_model_audit.csv`.

A negative allocation is an adjusted partitioning result, often associated
with shared information, suppression, collinearity, weak information, or
sampling variation. It is **not** evidence of a negative ecological effect.

For spatial SPD, the signed pooled allocation is
`r fmt_pct(value_for(spatial_spd, "signed", "climate"))` for climate and
`r fmt_pct(value_for(spatial_spd, "signed", "human"))` for human predictors.
The exact-zero sensitivity gives
`r fmt_pct(value_for(spatial_spd, "zero_truncated", "climate"))` climate,
while excluding complete models with any negative contribution gives
`r fmt_pct(value_for(spatial_spd, "exclude_negative", "climate"))` climate.
Across all five analyses, climate remains larger than human under all three
profiles: `r ifelse(all_conclusions_stable, "yes", "no")`.

```{r}
#| label: tbl-hvarpart-audit
#| tbl-cap: "Model-level HVarPart eligibility audit. Counts are not duplicated across predictor rows."
audit_overall |>
  select(
    analysis, n_models, n_eligible, n_excluded,
    n_negative_unique, n_negative_individual
  ) |>
  knitr::kable()
```

```{r}
#| label: fig-hvarpart-profiles
#| fig-cap: "Signed primary allocation and two sensitivity profiles. Values are pooled from unmodified fitted HVarPart results."
#| out-width: "100%"
knitr::include_graphics(
  file.path(figure_dir, "hvarpart_profile_comparison.png")
)
```

## Figure changes

Figure 2 has first been recreated with its region-map and predictor-panel
composition retained. It now labels the estimand "Signed allocation of
adjusted explained variation". The central view uses shared global 1st--99th
percentile limits rounded outward; boundary symbols and counts identify values
outside those limits. No value is removed from the exported data, and a
full-range supplementary rendering is provided.

Temporal H1 and H2 importance displays are unstacked and include explicit zero
and one reference lines. The established cores 14944 and 15394 show HVarPart's
original percentages and label total adjusted $R^2$ explicitly.

## Reconciliation

```{r}
#| label: tbl-hvarpart-validation
#| tbl-cap: "Programmatic reconciliation of formulas, exclusions, figure values, and artifacts."
validation |>
  knitr::kable()
```

All numerical statements in this response are computed from
`hvarpart_profile_comparison.csv`; audit statements use
`hvarpart_model_audit.csv`; and reconciliation uses
`hvarpart_validation_provenance.csv`.
